---
title: 'Google Data Analytics Specialization: Capstone Project - Cyclistic Bike-share'
author: "Adarsh K"
date: "17/08/2021"
output:
  word_document: default
  html_document: default
---

## Step 0: Importing Necessary Libraries
```{r Importing Libraries, warning=FALSE}
library(tidyverse, quietly = TRUE)
library(janitor, quietly = TRUE)
library(lubridate, quietly = TRUE)
library(ggplot2, quietly = TRUE)
#library(dplyr, quietly = TRUE)
```

***

## Step 1: Importing and Wrangling the Data
1. `starting_station_name` and `starting_station_id` are missing for some rows, ignore them.
2. Cyclistic collected some bikes to maintain them while these data were recorded. This is indicated with a negative or zero `trip_duration` value. Ignore them too.

```{r Importing Data}
df1 <- read.csv("202004-divvy-tripdata.csv")
df2 <- read.csv("202005-divvy-tripdata.csv")
df3 <- read.csv("202006-divvy-tripdata.csv")
df4 <- read.csv("202007-divvy-tripdata.csv")
df5 <- read.csv("202008-divvy-tripdata.csv")
df6 <- read.csv("202009-divvy-tripdata.csv")
df7 <- read.csv("202010-divvy-tripdata.csv")
df8 <- read.csv("202011-divvy-tripdata.csv")
df9 <- read.csv("202012-divvy-tripdata.csv")
df10 <- read.csv("202101-divvy-tripdata.csv")
df11 <- read.csv("202102-divvy-tripdata.csv")
df12 <- read.csv("202103-divvy-tripdata.csv")
```

```{r Combine the dataframes in to one big df}
bike_rides <- rbind(df1, df2, df3, df4, df5, df6, df7, df8, df9, df10, df11, df12)
```

```{r Data Cleaning as Mentioned in Points 1 and 2}
bike_rides <- janitor::remove_empty(bike_rides, which = c("cols"))
bike_rides <- janitor::remove_empty(bike_rides, which = c("rows"))
bike_rides <- bike_rides %>% 
                filter(start_station_name != "")
```

```{r Converting datetime to Date and Time}
bike_rides$Ymd <- as.Date(bike_rides$started_at)

bike_rides$started_at <- lubridate::ymd_hms(bike_rides$started_at)
bike_rides$ended_at <- lubridate::ymd_hms(bike_rides$ended_at)

bike_rides$start_hour <- lubridate::hour(bike_rides$started_at)
bike_rides$end_hour <- lubridate::hour(bike_rides$ended_at)
```

```{r Calculate the Difference in time (h and m)}
bike_rides$Hours <- difftime(bike_rides$ended_at, bike_rides$started_at, units = c("hours"))

bike_rides$Minutes <- difftime(bike_rides$ended_at, bike_rides$started_at, units = c("mins"))

bike_rides <- bike_rides %>% 
                filter(Minutes > 0)
```

***

## Step 2: Create a Summary df
```{r Summary df}
bike_rides2 <- bike_rides %>%
  group_by(weekly = floor_date(Ymd, "week"), start_hour) %>% 
  summarize(
    Minutes = sum(Minutes),
    Mean = mean(Minutes),
    Median = median(Minutes),
    Max = max(Minutes),
    Min = min(Minutes),
    Count = n()
  ) %>% ungroup()
```

### Summary of Hourly Counts
```{r Summary of Hourly Counts}
summary(bike_rides2$Count)
```

### Count of Rides by Hour
```{r Count of }
xtabs(bike_rides2$Count ~ bike_rides2$start_hour)
```

***

## Step 3: Plots
```{r Count Plot Per Day}
bike_rides2$Monthly <- lubridate::month(bike_rides2$weekly)

ggplot(data = bike_rides2) + geom_col(aes(x = weekly, y = Count)) + scale_y_continuous(labels = scales::comma) + labs(title = "Count of Rides per Day", subtitle = "Based on 28 day moving average", y = "Average Rides per Day", x = "") 
```

```{r Count Plot Per Hour}
ggplot(data = bike_rides2) + geom_col(aes(x = start_hour, y = Count)) + scale_y_continuous(labels = scales::comma) + labs(title = "Count of Rides per Hours", y = "Rides per Hour", x = "Ride Start Time") 
```

### Count of Rides by Bike Type (`rideable_type`)
#### Summary of Bike Types
```{r Summary of Bike Types}
bike_types <- bike_rides %>% 
  group_by(member_casual, rideable_type, weekly = floor_date(Ymd, "week")) %>% 
  summarize(
    Minutes = sum(Minutes),
    mean = mean(Minutes),
    Median = median(Minutes),
    Max = max(Minutes),
    Min = min(Minutes),
    Count = n()
  ) %>% ungroup()
```

#### Count by Bike Type (Total by Week)
```{r}
table(bike_types$rideable_type)
```
```{r Count of Rides by Bike Type}
ggplot(data = bike_types) + geom_col(aes(x = weekly, y = Count, fill = rideable_type)) + scale_y_continuous(labels = scales::comma) + labs(title = 'Count of Rides by Bike Type', subtitle = 'For 12 Months: Apr 2020 - Mar 2021', fill = "Bike Type", x = "")
```

```{r Top 10 Start Stations by Ride Count}
bike_rides %>% count(start_station_name, sort = TRUE) %>% top_n(10) %>% 
  ggplot() + geom_col(aes(x=reorder(start_station_name, n), y=n)) + coord_flip() + labs(x = "Station Name", y = "Count of Rides", title = "Top 10 Start Stations by Ride Count") 
```


```{r Count of Rides by Rider Type}
ggplot(bike_types) + geom_col(aes(x = weekly, y = Count, fill = member_casual)) + scale_y_continuous(labels = scales::comma) + labs(title = 'Count of Rides by Rider Type', subtitle = "For 12 Months: Apr 2020 - Mar 2021", fill = "Rider Type", x = "")
```

```{r Total Ride in Minutes by Week}
ggplot(bike_types) + geom_col(aes(x = weekly, y = Minutes)) + scale_y_continuous(labels = scales::comma) + facet_wrap(~rideable_type) + labs(title = "Total Ride in Minutes by Week", x = "") + theme(axis.text.x = element_text(angle = 45))
```

```{r Weekly Rides in Minutes by Bike Type}
ggplot(bike_types) + geom_col(aes(x = weekly, y = Minutes, fill = rideable_type)) + scale_y_continuous(labels = scales::comma) + labs(title = "Weekly Rides in Minutes by Bike Type", x = "", y = "Bike Trip in Minutes", fill = "Bike Type") 
```

***